Lactate transport inhibition therapeutically reprograms fibroblast

### 2024-03-11 manuscript/manuscript.qmd
finish Discussion first draft

Toward this end, we sought to characterize the effects of a novel metabolic approach - inhibition of lactate transport - on myofibroblast differentiation and experimental pulmonary fibrosis. Lactate must be secreted to sustain glycolysis in myofibroblasts. A family of monocarboxylate transporters (MCT1-4) maintains lactate homeostasis and small molecule inhibitors of these transporters have been under active clinical development for oncologic applications where glycolytic reprogramming also features prominently in disease pathobiology [@puri2020; @cluntun2021]. Importantly, compared to previously investigated inhibitors of glycolysis, MCT inhibitors exhibit favorable pharmacologic profiles and have successfully translated to human clinic trials [@halford2023]. Before the promise of this therapeutic approach in IPF may be realized, however, the preclinical efficacy and molecular mechanisms-of-action of lactate transport inhibitors must be demonstrated experimentally.

In this work, we evaluated the contribution of lactate transporters to experimental pulmonary fibrosis. We found increased expression of the lactate transporters MCT1 and MCT4 IPF patient lungs, consistent with increased glycolysis previously observed. Inhibition of these transporters attenuated bleomycin-induced lung fibrosis *in vivo* and TGFβ-induced myofibroblast differentiation *in vitro*, where MCT4 inhibition demonstrated increased therapeutic efficacy. *Using stable isotope tracing, high-resolution spatial metabolomics, we find...* Based on these findings, we report a novel MCT4 inhibitor suitable for human clinical studies. Together, these data illuminate a novel approach to disrupt the metabolic program fueling the exuberant fibrosis at the core of IPF and provide proof-of-concept to target lactate transporters in the treatment of this deadly disease.

@@

Lactate metabolism is closely coupled to cellular redox homeostasis through its metabolism by lactate dehydrogenases, which transfer electrons from lactate to NADH. Since prior studies have suggested that reactive oxygen species (ROS) are important for TGFβ-mediated gene expression [@jain2013, @schworer2020], we next examined the impact of lactate transport inhibition on cellular redox homeostasis. Consistent with an increase in intracellular lactate with MCT4 inhibition, we observed a concomitant increase in intracellular NADH/NAD^+^ (**Figure 7A**), which was associated with a reduction in total ROS as measured by the CellROX fluorescent probe (**Figure 7B**). By contrast, the NADPH/NADP^+^ was unchanged by either TGFβ or MCT inhibitors (**Supplemental Figure 7A**). Notably, we did not observe TGFβ-dependent increases in total ROS production nor an impact of TGFβ or inhibitors on mitochondrial superoxide production (**Supplemental Figure 7B**).

Proline biosynthesis has been previously suggested as a mechanism by which NIH-3T3 fibroblasts prevent excessive ROS accumulation following TGFβ stimulation [@schworer2020]. In this model, proline synthesis from glutamine consumes reducing equivalents from NADPH and NADH, thereby ameliorating reductive stress and decreasing ROS production. Compared to this prior study, we did not find that TGFβ increased proline in our primary lung fibroblasts (**Supplemental Figure 7C**). Overall, the fractional labeling of proline from [U-^13^C~5~]-glutamine was modest (~10%) compared to the 40% previously reported (**Supplemental Figure 7D**). While we observed trends toward increased proline production from glutamine with MCT4 inhibition, this mechanism seems unlikely to account for the effects of MCT inhibitors on ROS generation.

@@

Compared to young mice, bleomycin-induced fibrosis in aged mice is more severe and does not resolve spontaneously [@redente2011]. Given that IPF incidence increases with age, some studies have suggested such aged mouse models may be more clinically relevant. Thus, we examined the antifibrotic effects of VB253 in aged (60+ weeks) mice. Similar to young mice, VB253 decreased fibrosis severity as quantified by Ashcroft score and α-SMA expression (**Figure 10H-J**). The magnitude of improvement was similar to that observed with nintedanib and pirfenidone. As anticipated, VB2253 decreased total lung lactate similar to VB124 (**Figure 9A**). Taken together, these data provide compelling preclinical data for the development of lactate transporter inhibition as a therapeutic strategy for fibrotic lung disease.

@@

Craig Thompson paper: Lactate activates ETC independently of its metabolism

Craig Thompson paper: Isotope labeling in lung fibroblasts

<!-- spatial metabolomics -->

>>>>

Toward this end, we sought to characterize the effects of a novel metabolic approach - inhibition of lactate transport - on myofibroblast differentiation and experimental pulmonary fibrosis. Lactate must be secreted to sustain glycolysis in myofibroblasts. A family of monocarboxylate transporters (MCT1-4) maintains lactate homeostasis and small molecule inhibitors of these transporters have been under active clinical development for oncologic applications where glycolytic reprogramming also features prominently in disease pathobiology [@puri2020; @cluntun2021]. Importantly, compared to previously investigated inhibitors of glycolysis, MCT inhibitors exhibit favorable pharmacologic profiles and have successfully translated to human clinical trials [@halford2023]. Before the promise of this therapeutic approach in IPF may be realized, however, the preclinical efficacy and molecular mechanisms-of-action of lactate transport inhibitors must be demonstrated experimentally.

In this work, we evaluated the contribution of lactate transporters to experimental pulmonary fibrosis. We found increased expression of the lactate transporters MCT1 and MCT4 IPF patient lungs, consistent with increased glycolysis previously observed. Inhibition of these transporters attenuated bleomycin-induced lung fibrosis *in vivo* and TGFβ-induced myofibroblast differentiation *in vitro*, where MCT4 inhibition demonstrated increased therapeutic efficacy. Using metabolomics, stable isotope tracing, and high-resolution spatial metabolomic imaging, we find that lactate transport inhibition promotes glucose oxidation. These metabolic changes signal, in part, through increased protein lactylation and decreased intracellular reactive oxygen species (ROS) production. Based on these findings, we report a novel MCT4 inhibitor suitable for human clinical studies. Together, these data illuminate a novel approach to disrupt the metabolic program fueling the exuberant fibrosis at the core of IPF and provide proof-of-concept to target lactate transporters in the treatment of this deadly disease.

@@

Lactate metabolism is closely coupled to cellular redox homeostasis through its metabolism by lactate dehydrogenases, which transfer electrons from lactate to NADH. Since prior studies have suggested that reactive oxygen species (ROS) are important for TGFβ-mediated gene expression [@jain2013; @schworer2020], we next examined the impact of lactate transport inhibition on cellular redox homeostasis. Consistent with an increase in intracellular lactate with MCT4 inhibition, we observed a concomitant increase in intracellular NADH/NAD^+^ (**Figure 7A**), which was associated with a reduction in total ROS as measured by the CellROX fluorescent probe (**Figure 7B**). AZD3965, but not VB124, also decreased mitochondrial superoxide production (**Supplemental Figure 7B**). By contrast, the NADPH/NADP^+^ was unchanged by either TGFβ or MCT inhibitors (**Supplemental Figure 7A**). Notably, we did not observe TGFβ-dependent increases in total ROS production in our model system.

Proline biosynthesis has been previously suggested as a mechanism by which NIH-3T3 fibroblasts prevent excessive ROS accumulation following TGFβ stimulation [@schworer2020]. In this model, proline synthesis from glutamine consumes reducing equivalents from NADPH and NADH, thereby ameliorating reductive stress and decreasing ROS production. Compared to this prior study, we did not find that TGFβ increased proline in primary lung fibroblasts (**Supplemental Figure 7C**). Overall, the fractional labeling of proline from [U-^13^C~5~]-glutamine was modest at 10% compared to the 40% previously reported (**Supplemental Figure 7D**). While we observed trends toward increased proline production from glutamine with MCT4 inhibition, this mechanism seems unlikely to account for the effects of MCT inhibitors on ROS generation.

@@

Compared to young mice, bleomycin-induced fibrosis in aged mice is more severe and does not resolve spontaneously [@redente2011]. Given that IPF incidence increases with age, some studies have suggested such aged mouse models may be more clinically relevant. Thus, we examined the antifibrotic effects of VB253 in aged (60+ weeks) mice. Similar to young mice, VB253 decreased fibrosis severity as quantified by Ashcroft score and α-SMA expression (**Figure 10H-J**). The magnitude of improvement was similar to that observed with nintedanib and pirfenidone. As anticipated, VB253 decreased total lung lactate similar to VB124 (**Figure 9A**). Taken together, these data provide compelling preclinical data for the development of lactate transporter inhibition as a therapeutic strategy for fibrotic lung disease.

@@

Our findings demonstrate that lactate transport is an essential component of the metabolic reprogramming associated with myofibroblast differentiation *in vitro* and *in vivo*. We observed increased expression of the lactate transporters MCT1 and MCT4 in IPF lung explants and experimental model systems. Inhibition of these transporters decreased TGFβ-stimulated myofibroblast differentiation and the severity of bleomcyin-induced pulmonary fibrosis. Metabolically, MCT antagonists promoted glucose oxidation and decreased glucose carbon incorporation into fibrotic lung regions, which were associated with decreased oxidative stress. Beyond metabolic reprogramming, we observed changes in gene expression profiles consistent with antifibrotic metabolic signaling downstream of lactate transport inhibition, which may be mediated by changes in ROS production, protein lactylation, or other mechanisms. MCT4 inhibition consistently proved more potently antifibrotic than MCT1 inhibition, and we report a novel MCT4 inhibitor, VB253, with a more favorable pharmacologic profile than VB124 that is now in Phase 1 clinical trials [**REF?**]. Collectively, these data establish lactate transport as a metabolic opportunity for therapeutic intervention in pulmonary fibrosis.

Metabolic reprogramming is a hallmark of myofibroblast differentiation. Previous studies have identified alterations in carbohydrate, amino acid, and lipid metabolic pathways that support fibrogenesis [@selvarajah2021]. Of these, increased lactate production was among the earliest recognized metabolic changes associated with pulmonary fibrosis [@kottmann2012]. Subsequent work demonstrated increased lactate production was driven by activation of glycolysis in myofibroblasts [@xie2015] and that inhibition of glycolysis prevented myofibroblast differentiation [@xie2015; @kottmann2015; @bernard2015] and, in turn, experimental pulmonary fibrosis [@judge2018; @cho2017]. Consistent with increased glycolysis in myofibroblasts, we have now demonstrated increased expression of the lactate transporters MCT1 and MCT4 support this metabolic phenotype.

MCTs are proton-coupled monocarboxylate symporters with varying affinities for lactate, pyruvate, and other monocarboxylates [@halestrap2013]. MCT1 is constitutively and ubiquitously expressed and primarily considered to be a lactate importer with a K~M~ for lactate of 3-6 mM. By contrast, MCT4 expression is dynamically regulated, including by HIF-1α, and has primarily been considered to be a lactate exporter with a low lactate affinity (K~M~ 30-40 mM). Recent data suggests, however, that MCT4 has a much higher affinity for lactate import than previously appreciated (K~M~ 1 mM) [@contreras-baeza2019]. These data are most consistent with our findings that inhibition of both MCT1 and MCT4 was required to inhibit lactate export (**Figure 3**) and isotope incorporation from extracellular [U-^13^C~3~]-lactate import (**Figure 4J**). Previous studies of cancer cells have similarly recognized that MCT4-expressing cells are resistant to the cytotoxic effects of MCT1 inhibition [@curtis2017]. Interestingly, inhibition of either MCT1 or MCT4 reduced myofibroblast differentiation without affecting lactate export.

To clarify the metabolic consequences of lactate transporter inhibition, we performed a comprehensive metabolic analysis of myofibroblasts treated with MCT inhibitors, including bioenergetic measurements, metabolomic profiling, and stable isotope tracing experiments. The primary metabolic consequence of MCT inhibition is activation of oxidative phosphorylation. MCT inhibitors increased the fraction of mitochondrial ATP production, which was associated with increased levels of tricarboxylic acid cycle intermediates, and increased isotope incorporation from glucose substrates. Collectively, these data indicate that MCT inhibition reroutes carbon flow from glucose away from lactate fermentation and toward glucose oxidation. Several mechanisms may account for this phenomenon. Since the magnitude of these effects generally correlates with intracellular lactate levels, mass action may be an important driver of this metabolic phenomenon. Lactate accumulation is also closely coupled in NADH production through lactate dehydrogenase activity. Cytoplasmic NADH may be transported into the mitochondria through the malate-aspartate shuttle and oxidized by the electron transport chain. Recently, lactate itself was shown activate the mitochondrial electron transport chain independently of its metabolism [@cai2023], although the molecular mechanism for this effect is unknown.

How these metabolic effects signal to antifibrotic transcriptional programs remains unclear. We investigated several hypotheses based on previous studies. First, we did not observe significant changes in canonical (*i.e.*, Smad3) or non-canonical (*i.e.*, ERK) TGFβ signaling pathways. Second, TGFβ-dependent activation of HIF-1α seems to be required for myofibroblast differentiation [@goodwin2018]. MCT1 and MCT4 potentiated HIF-1α stabilization following TGFβ (**Figure 6D**), indicating that MCT inhibition acts downstream of HIF signaling pathways. Third, post-translational modification of histone lysines by lactate serves as an epigenetic modification that stimulates gene expression [@zhang2019a]. We observed increased histone lactylation following dual MCT inhibition, which may account for some of the transcriptional changes we observed. In addition to histones, protein lactylation has been increasingly identified as a critical regulator of protein function, including cytoskeletal proteins [@yang2023]. Fourth, increased ROS production has been observed following TGFβ treatment [@jain2013; @schworer2020]. While we were unable to replicate TGFβ-dependent increases in ROS in our cells and model system, we did observe decreased CellROX oxidation with MCT4 inhibition and decreased MitoSOX oxidation with AZD3965. Collectively, these studies identify two putative mechanisms for antifibrotic lactate signaling following MCT inhibition, protein lactylation and antioxidant activity; however, novel lactate targets are being increasingly recognized as important mediators of metabolic signaling [@liu2023; @liu2023b].

Through our detailed metabolic investigation of lactate transport inhibition, we have also generated valuable data on the metabolic and transcriptional consequences of human lung myofibroblast differentation. TGFβ stimulated marked changes in amino acid and nucleic acid metabolic pathways (**Supplemental Figure 4A-D**). Previous studies have identified important roles for glutamine, proline, and taurine [@selvarajah2021; @choudhury2023; @phan1995], while the contributions of branched chain amino acid metabolism, for example, remain unexplored. Integrating these multi-omics data sets may identify novel molecular targets for future drug development.

In addition to metabolomic profiling, we have performed stable isotope tracing using glucose, lactate, and glutamine substrates in TGFβ-treated primary human lung fibroblasts. To our knowledge, these are the first comprehensive data on intracellular substrate metabolism in this commonly utilized model of myofibroblast differentiation. Overall, isotope labeling patterns changed little following TGFβ treatment (**Supplemental Figure 4.2**). Together with the results of our extracellular flux experiments, this finding suggests that TGFβ primarily increases the flow of metabolites through metabolic pathways without substantially altering the pathways themselves. Consistent with increased glucose uptake and accumulation of glycolytic metabolites, a greater fraction of pyruvate, lactate, alanine, and serine are labeled by [U-^13^C~6~]-glucose, which is offset by a decrease in the fractional labeling from [U-^13^C~3~]-lactate. Similar to our previous work [@copeland2023], nearly 50% of tricarboxylic acid metabolites (citrate, 2-oxoglutarate, succinate, and malate) are labeled by [U-^13^C~3~]-lactate, highlighting the importance of lactate as a respiratory fuel source in these cells.

In addition to cellular bioenergetics, the contribution of fibroblast metabolic pathways to extracellular matrix production has been an intense area of investigation [@selvarajah2021]. Serine and glycine may be synthesized from the glycolytic intermediate 3-phosphoglycerate, and proline may be synthesized from glutamine. These pathways have been implicated in myofibroblast differentation and pulmonary fibrosis [@bernard2018; @hamanaka2019; @selvarajah2019; @schworer2020; @choudhury2023]. Interestingly, our data suggest little incorporation of glucose or lactate carbon into serine and little incorporation of glutamine carbon into proline. These differences may be secondary to the cells studied (IMR-90 fetal lung fibroblasts or NIH-3T3 spontaneously immortalized mouse embryonic fibroblasts) or the culture medium utilized (Eagle's minimum essential medium, which lacks serine, glycine, and proline). In this context, our study highlights that fibroblasts will preferentially use available amino acids rather than reroute substrates into biosynthetic pathways. Developing metabolic flux models using human physiologic medium [@cantor2017] may provide a more accurate and complete characterization of substrate flow into energetic and biosynthetic pathways that more closely reflects fibroblast metabolism *in vivo*.

Even better than the use of *in vitro* model systems are approaches that enable the study of cell metabolism *in vivo*. Here, we performed metabolomic profiling of lung and plasma samples from mice treated with bleomycin and MCT inhibitors (**Supplemental Figure 9**). These studies did not reveal consistent changes in whole lung metabolite levels and demonstrate a potential limitation of using bulk metabolomics to monitor metabolic changes in heterogeneous cell populations *in vivo*. We sought to overcome this limitation with spatial metabolic imaging. Mice received stable isotope tracers of proline and glucose prior to euthanasia. Since these tracers are administered to live animals, multi-isotope imaging mass spectrometry identifies lung regions that are metabolically active over the labeling period. Since tissues are fixed and processed *ex vivo*, the isotope signals indicate substrate incorporation into fixable biomass [@zhang2020; @wertheim2023]. Thus, using multi-isotope imaging mass spectrometry, we are directly quantifying isotope flux from glucose into fibrotic lung regions. We found that ^15^N-proline enrichment levels correlate with tissue fibrosis from histologic and biochemical assessments. This finding emphasizes the ability of cells to incorporate circulating proline for protein synthesis, and likely de-emphasizes the importance of *de novo* proline biosynthesis *in vivo*. The ^2^H-glucose signal provides some direct evidence for MCT-dependent metabolic reprogramming in mice, with less carbohydrate incorporation into matrix proteins. Owing to the financial and time costs of isotope tracing and imaging, only a few animals per group were studied. Nonetheless, metabolic imaging at high spatial resolution offers tremendous promise for correlating cell identities determined from spatial transcriptomic profiles with metabolic features. Stable isotopes may be safely administered to human patients prior lung biopsy or explant, an approach that would considerably advance our understanding of how cell metabolism contributes to pulmonary fibrosis.

MCT1 and MCT4 are expressed by many cells in the lung, particularly macrophages and dendritic cells [@travaglini2020]. Although our *ex vivo* and *in vitro* experiments suggest that inhibiting myofibroblast differentiation is a primary antifibrotic mechanism, inhibition of lactate transporters expressed by other cell types may also play a role in their therapeutic effect. For example, MCT4 expression is increased as part of a HIF-1α gene expression signature in transitional AT2 cells that accumulate in pulmonary fibrosis and contribute to aberrant repair [@choi2020; @zhu2022]. We are in the process of generating fibroblast-specific conditional knockout mice to explore the role of these transporters in fibroblasts specifically.

Pharmacologic interest in lactate transporters has been driven by the recognition of increased lactate transporter expression in a variety of cancers [@payen2020]. AZD3965 was selected for this study as it has been investigated in human clinical trials for advanced solid organ malignancies [@halford2023]. In this Phase 1 study, AZD3965 was generally well-tolerated with 7 of 40 patients experiencing dose-limiting toxicities including asymptomatic, reversible ocular changes; acidosis; and increased troponin. VB124 was the first selective MCT4 inhibitor developed [@cluntun2021] and here we report a second generation MCT4 inhibitor, VB253. A Phase 1 clinical trial of VB253 is currently enrolling [**REF?**]. MCT4 global knockout mice are viable and breed normally [@bisetto2018; @bisetto2019], increasing optimism that VB253 will be well tolerated in humans. One of the barriers to translating metabolic therapies to humans has been the poor pharmacologic properties of the small molecule laboratory compounds that have been studied previously. The IC~50~ values for AZD3965 and VB253 are ~ 2 nM, compared to the next most potent inhibitor, lonidamine, with an IC~50~ of 7,000 nM [@yin2019]. As a metabolic target downstream of glycolysis, lactate transport inhibitors may also be better tolerated by allowing glycolysis to continue to support of glucose oxidation, while upstream inhibitors of glycolysis have a more severe impact on cellular bioenergetics.

In conclusion, we have demonstrated that lactate transporters promote myofibroblast differentiation and pulmonary fibrosis. Using a comprehensive metabolic phenotyping strategy, we have identified antifibrotic mechanisms for MCT inhibition and demonstrated evidence for metabolic reprogramming in animal models. We have demonstrated the antifibrotic efficacy of currently available lactate transport inhibitors with accepted preclinical disease models. Collectively, our data advance lactate transport inhibition as a viable therapeutic strategy for patients with pulmonary fibrosis.
